---
title: "Motor Trends mtcars analysis"
author: "Alejandro Borges Sanchez"
date: "7/12/2021"
header-includes:
  - \usepackage{mathtools}
  - \usepackage{fontspec}
  - \setmainfont{Arial}
fontsize: 10pt
output:
  pdf_document: 
    latex_engine: xelatex
  html_document:
    df_print: paged
  html_notebook: default
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE,
	fig.width = 6,
	fig.height = 6
)
library(tidyverse)
library(DataExplorer)
library(DT)
data("mtcars")
library(GGally)
library(ggfortify)
library(magrittr)
library(broom)
thecars<-mtcars%>%
    rename(transmission=am)%>%
    mutate(transmission=if_else(transmission==0,"A","M"),
           transmission=as_factor(transmission),
           wt=wt*1000,
           vs=if_else(vs==0,"V","S"),
           p2w=hp/wt,#I converted wt to pounds back at the beginning: yields hp/lb
         power_loading=1/p2w, #the inverse of power to weight is easyer to think about, the bigger, the heavyer and/or less hp, the smaller, the lighter and/or more powerful
         power_loading=power_loading-mean(power_loading))%>% #Better if centered)
  as_tibble(rownames = "Model")
ggplot2::theme_set(ggplot2::theme_classic())
```

# Executive Summary

A list of 32 cars have been submitted to me for examination of the relationship
between Miles Per Gallon and the transmission type. I am asked:

-   Is automatic transmission better than manual transmission?
-   Can we quantify this difference?

A set of models where attempted on the data set taking account each and all
variables. Each variable was explored in-depth both as mathematical quantities
and as physics aspects of a car design, including its engine, gasoline
displacement and other aspects, not explicitly included in the original dataset,
such as power-to-weight ratio, were calculated and explored.

Power loading is the most important variable of this result, so its important to
define it explicitly:

$$
w2p=power\_loading=Weight/HorsePower
$$

Refer to the variable table in de appendix for further information this and
other measures.

## Preliminary analysis

The first attempt is to compare the difference in the average cars in both
manual and automatic cars, assuming this dataset represents the whole car
industry, that is, this sample of 32 cars is randomly selected and sufficiently
large, these are the results:

-   Cars with manual tranmission range in consumption from 21.6 to 27.16 miles
    per gallon while the average automatic transmission car ranges from 10.8 to
    23.5 miles per gallon, with 95% confidence.
-   As for the "can we quantify this", the answer is partially yes: most
    automatic and manual cars lie within that range, but its obvious that some
    automatic cars give more miles per gallon than manual cars and viceversa:
    some manual cars give less miles per gallon than some automatic cars.
    Speciffically, in this dataset, this cars are:

```{r fig.cap="This comes from the basic linear model where lm(mpg~transmission)"}
thecars%>%
  mutate(row_number=row_number())%>%
  filter(between(mpg,21.6,23.6))%>%
  select(row_number,Model,mpg,transmission,power_loading,carb)%>%
  knitr::kable()
```

Which you can immediatly see give exactly the same MPG, but are of auto and
manual transmissions. If these two cars represent whole classes of cars in the
market, then its possible that attempting to explain MPG only by transmission is
misguided. A full exploration of the many possible other variables in the
dataset that might help explain MPG additionally to transmission is available in
the appendix. Here, I list only the final conclussion:

## Conclusion 

After a length exploration of the data and several models, I arrived at this
conclusion:

-   Many factors other than transmission affect MPG. It is a more complex
    question than what one would think.

-   The main factor I included is the weight-to-power ratio, also called the
    power loading, which divides the car's weight in pounds over the cars horse
    powers.

-   Therefore, taking into account the power loading and the transmission type,
    I have 95% confidence that:

    -   Manual cars:

        -   The average power loading manual transmission car will travel 22.7
            to 27.2 miles per gallon.

    -   Automatic cars:

        -   The average power loading automatic transmission car will travel
            11.5 to 21.9 miles per gallon.

This does not tell the whole story. For each 1 unit of
$$w2p=power\_loading=Weight/HorsePower$$ ratio that the car goes over the
average power loading, any of this cars will travel from 0.2 to 0.57 additional
miles per gallon spent.

This might seem anti climatic for a magazzine, but please refer to the figures
in the appendix for a suggestion on how to publish this results. Cars can be
classified as better or worse in power loading and that position of the cars
with respect to power loading and mpg is what tells the rest of the story.

Due to this document's size constraint, I cannot put here all the exploration
necessary. Please refer to the github repo for [all the glorious
analysis](https://github.com/elfatherbrown/motor_trend) you might wish to see.

Suffice it to say that ANOVA suggests power loading is a great variable to
include and it doesnt have much impact in variance inflation:

```{r}
fit0<-lm(mpg~transmission,thecars)
fit1<-lm(mpg~transmission+power_loading,thecars)
print("ANOVA")
anova(fit0,fit1)
print("Variance Inflation Factor")
car::vif(fit1)
```

## Appendix

```{r fig.cap="How i would explain this results to people: cuadrantize both groups by mean mpg and mean power loading"}

meanmpgs<-thecars%>%
group_by(transmission)%>%
summarize(mean_mpg=mean(mpg),
          mean_power_loading=mean(1/p2w)
)

thecars%>%
ggplot(aes(x=1/p2w,y=mpg,label=Model))+
geom_point()+
ggrepel::geom_label_repel(size=2)+
geom_vline(data=meanmpgs,aes(xintercept=mean_power_loading))+
geom_hline(data=meanmpgs,aes(yintercept=mean_mpg))+
facet_wrap(transmission~.)

```

### Other explorations

```{r fig.height=6, fig.width=8, message=FALSE, warning=FALSE}

ggpairs(data = thecars%>%select(-Model),
        ggplot2::aes(colour=transmission),
        upper=list(continuous=wrap("cor",size=2)),
        diag=list(continuous=wrap("diagAxis",labelSize=2,gridLabelSize=2))
        )+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 0))


```

+---------+-----------------------+--------------------------+-----------+---------------+
| Var     | Doc                   | Note                     | Source    | f(?)          |
+=========+=======================+==========================+===========+===============+
| mpg     | M iles/(US) gallon    |                          |           |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| cyl     | Number of cylinders   |                          |           |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| disp    | Dis placement         | How much gas/air mix     | \<https   | f(cyl,volcyl  |
|         | (cu.in.)              | passes through the       | ://       | )=cyl\*VolCyl |
|         |                       | cylinders in total       | www.yourm |               |
|         |                       |                          | ech       |               |
|         |                       |                          | anic.com/ |               |
|         |                       |                          | art       |               |
|         |                       |                          | icle/what |               |
|         |                       |                          | -is       |               |
|         |                       |                          | -engine-d |               |
|         |                       |                          | ispl      |               |
|         |                       |                          | acement\> |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| hp      | Gross h orsepower     |                          | \<<h      | f(Torque,     |
|         |                       |                          | ttps://w> | RPM)=T\*RPM+e |
|         |                       |                          | ww.       |               |
|         |                       |                          | wikihow.c |               |
|         |                       |                          | om/       |               |
|         |                       |                          | Calculate |               |
|         |                       |                          | -Hor      |               |
|         |                       |                          | sepower\> |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| drat    | Rear axle ratio       | Motor revs/rear axle     | \<        | f(RPMaxlefron |
|         |                       | revs ( kind of a measure | <https:/> | t,RPMaxleback |
|         |                       | of efficiency )          | /ww       | )=RPMaxelfron |
|         |                       |                          | w.autolis | t/RPMaxleback |
|         |                       |                          | t.c       |               |
|         |                       |                          | om/guides |               |
|         |                       |                          | /axl      |               |
|         |                       |                          | e-ratio\> |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| wt      | Weight (1000 lbs)     |                          |           | f(car)        |
+---------+-----------------------+--------------------------+-----------+---------------+
| qsec    | 1/4 mile time         |                          |           |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| vs      | Engine (0             | V: classic, can have     | \<ht      | f(cyl)        |
|         |                       | more cylinders,          | tp:       |               |
|         | V-shaped, 1 =         | Straight: more           | //www.why |               |
|         | straight)             | efficient, normally less | hig       |               |
|         |                       | cylinders                | hend.com/ |               |
|         |                       |                          | inl       |               |
|         |                       |                          | ine-vs-v- |               |
|         |                       |                          | engi      |               |
|         |                       |                          | ne.html\> |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| am      | Tra nsmission (0 = a  |                          |           |               |
|         | utomatic, 1 = manual) |                          |           |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| gear    | Number of forward     |                          |           |               |
|         | gears                 |                          |           |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| carb    | Number of ca          |                          |           |               |
|         | rburetors             |                          |           |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| p2w     | hp/(wt\*1000)         |                          |           |               |
+---------+-----------------------+--------------------------+-----------+---------------+
| power_  | 1/p2w-(mean(1/p2w))   | [https://en              |           |               |
| loading |                       | .wikipedia.org/wiki/Powe |           |               |
|         |                       | r-to-weight_ratio](https |           |               |
|         |                       | ://en.wikipedia.org/wiki |           |               |
|         |                       | /Power-to-weight_ratio)) |           |               |
+---------+-----------------------+--------------------------+-----------+---------------+

: The variables themselves, with research links

```{r fig.cap="Model:lm(mpg~transmission,thecars)"}
lm(mpg~transmission,thecars)%>%autoplot()
```

```{r fig.cap="Model:lm(mpg~transmission+power_loading,thecars)"}
lm(mpg~transmission+power_loading,thecars)%>%autoplot()
```
