---
title: "Best MPG by transmission type"
author: "Alejandro Borges Sanchez"
date: "7/12/2021"
header-includes:
  - \usepackage{mathtools}
  - \usepackage{fontspec}
  - \setmainfont{Arial}
fontsize: 10pt
output:
  pdf_document: 
    latex_engine: xelatex
  html_document:
    df_print: paged
  html_notebook: default
editor_options:
  markdown:
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE,
	fig.width = 4,
	fig.height = 4
)
library(tidyverse)
library(DataExplorer)
library(DT)
data("mtcars")
library(GGally)
library(ggfortify)
library(magrittr)
library(broom)
thecars<-mtcars%>%
    rename(transmission=am)%>%
    mutate(transmission=if_else(transmission==0,"A","M"),
           transmission=as_factor(transmission),
           wt=wt*1000,
           vs=if_else(vs==0,"V","S"))%>%
  as_tibble(rownames = "Model")
ggplot2::theme_set(ggthemes::theme_solarized_2())
```

## The questions

BOSS: Our magazzine is to finally settle the debate once and for all! Alex! Get
on to work on this list of cars here and answer NOW:

-   "Is an automatic or manual transmission better for MPG"
-   "Quantify the MPG difference between automatic and manual transmissions"

### Investigation 1

Okay boss. As you know I'm not that much into the intricate measures of our
cars. My stuff has to do more with the now (70's) booming electronic tech side
of automotive stuff.

So I took all the data you gave me and did some research in the internet (which
does not yet exist, but okay). This yielded:

+-------+--------------+-------------------------------+-----------------------+
| Var   | Doc          | Note                          | Source                |
+-------+--------------+-------------------------------+-----------------------+
| mpg   | Miles/(US)   |                               |                       |
|       | gallon       |                               |                       |
+-------+--------------+-------------------------------+-----------------------+
| cyl   | Number of    |                               |                       |
|       | cylinders    |                               |                       |
+-------+--------------+-------------------------------+-----------------------+
| disp  | Displacement | How much gas passes through   | \<                    |
|       | (cu.in.)     | the cylinders                 | <https://www.yourmec> |
|       |              |                               | ha                    |
|       |              |                               | nic.com/article/what- |
|       |              |                               | is-                   |
|       |              |                               | engine-displacement\> |
+-------+--------------+-------------------------------+-----------------------+
| hp    | Gross        |                               |                       |
|       | horsepower   |                               |                       |
+-------+--------------+-------------------------------+-----------------------+
| drat  | Rear axle    | Motor revs vs. rear axle revs | \<<                   |
|       | ratio        |                               | https://www.autolist> |
|       |              |                               | .co                   |
|       |              |                               | m/guides/axle-ratio\> |
+-------+--------------+-------------------------------+-----------------------+
| wt    | Weight (1000 |                               |                       |
|       | lbs)         |                               |                       |
+-------+--------------+-------------------------------+-----------------------+
| qsec  | 1/4 mile     |                               |                      |
|       | time         |                               |                       |
+-------+--------------+-------------------------------+-----------------------+
| vs    | Engine (0 =  | V: classic, can have more     | \<http                |
|       | V-shaped, 1  | cylinders, Straight: more     | :/                    |
|       | = straight)  | efficient, normally less      | /www.whyhighend.com/i |
|       |              | cylinders                     | nli                   |
|       |              |                               | ne-vs-v-engine.html\> |
+-------+--------------+-------------------------------+-----------------------+
| am    | Transmission |                               |                       |
|       | (0 =         |                               |                       |
|       | automatic, 1 |                               |                       |
|       | = manual)    |                               |                       |
+-------+--------------+-------------------------------+-----------------------+
| gear  | Number of    |                               |                       |
|       | forward      |                               |                       |
|       | gears        |                               |                       |
+-------+--------------+-------------------------------+-----------------------+
| carb  | Number of    |                               |                       |
|       | carburetors  |                               |                      |
+-------+--------------+-------------------------------+-----------------------+
|       |              |                               |                       |
+-------+--------------+-------------------------------+-----------------------+

An *excruciating* report is included in the appendix as to how and why did I
arrive at the following conclusions to settle this ages old debate!

### Discussion

ALEX: (after working a bit)

Okay boss, If you want a straight answer to the first and second questions I can
tell you right now:

-   Manual cars have 21.6 to 27.2 consumption on Average

-   Automatic cars have 10.8 to 23.56 consumption on Average

```{r}
tidy(lm(mpg~transmission,thecars),conf.int = TRUE)
```

So err... um... well for a bunch of cars automatic transmission reduces
efficiency (less miles per gallon spent), but for some of them it might be
um...about the same... :D

BOSS: Remind me why do we pay your very expensive salary, Alex? I need MORE than
that puny attempt at a guess!

ALEX: (sighs) (thinks: back to the mines, then)... Yes, boss...

(after a while)

Okay I GOT IT! The most important thing that can impact MPG in a vehicle cannot
be transmission on its own! We have different engine designs like V shaped vs.
straight, that allows for more or less cylinders. We have the weight of
vehicles, which impact how much gas demand is the motor going to experience. We
have also the number of cylinders themselves, of course, although that is no
good indicator by itself (for one, you might have many cylinders, but they
aren't the same size across vehicles!).

So we need to assume some things before we start, because you want to infer,
based on this data, if MPG of all cars depends on transmission and by how much.

-   So the first assumption is that the cars in the data are randomly selected.

So if that holds true, then everything else down there is the truth, the whole
truth and nothing but the truth!

-   The second assumption is that at the very least MPG depends on the
    transmission type (everybody knows that!, right?) and the engine design
    (vshaped vs. straight in line).

For the first assumption, you selected the data so only you know if its true.
But for the second, we can test! But first, lets assume its true. Check it out
boss:

```{r}
fit0<-lm(mpg~transmission+vs,thecars)
    tidy(fit0,conf.int = TRUE)
```

Ok BOSS!,I know you ain't a numbers guy, so let me translate to English. What
that is saying is:

"The average miles per galon of a Manual transmission car with a Straight engine
design is between 25.3 and 29.9 with 95% confidence"

Furthermore:

"Automatic transmission vehicles with a straight engine increase that average by
between 6.7 and 3.4 miles per gallon with 95% certainty"

If, however, the engine is V-shaped:

"For manual or automatic transmissions, vshaped engine designs present and
average consumption descent of 9.5 to 4.34 miles per galon when compared to
straight engines, which makes a lot of sense".

The most efficient car you can buy, IF it is true that MPG depends mostly or
only on engine shape and transmission, is Manual transmission with V shaped
engine.

## Appendix

### Modeling transmission as binomial response

I think we can do a bernoulli fit where the transmission is the output binary
variable and we can see how much does an increase or decrease of mpg influence
the probability that a car has automatic or manual transmission.

Remember then that, in this model, Manual is coded as 1 and Automatic is coded
as zero. Additionally, mpg as a covariate must be centered so that we might have
a chance to interpret the intercept as the probability for the cars with mean
mpg to have a manual transmission.

```{r}
# simplefit_binom_centered_mpg<-glm(formula = transmission ~ mpg, family = "binomial", 
#                                   data = thecars%>%mutate(mpg=mpg-mean(mpg))
#                                   )
# tibble(x=thecars$mpg,y=fitted.values(simplefit))%>%
# ggplot(aes(x=x,y=y))+
#     geom_line()+
#     geom_point(data=tibble(mtcars,transmission=thecars$transmission),aes(x=mpg,y=am,color=transmission))
```

Remember that Y is in the logit scale. We can see that our fitted model captures
a basic idea: if you increase mpg, the log odds of the cars having a manual
transmission decreases. We will transform this to the probability scale
($$p_i=exp(\beta_0+\beta_1)/(1+exp(\beta_0+\beta_1))$$) and remove the points:

```{r}
# tibble(x=thecars$mpg,y=exp(fitted.values(simplefit)+0.435)/(1+exp(fitted.values(simplefit)+0.435)))%>%
#     ggplot(aes(x=x,y=y))+
#     geom_line()
```

Our model is saying then that for lower miles per gallon, the probability of a
car having a manual transmission is much higher than for higher miles per
gallon.

Now we can dance around this all we want, but it means squat if we don't check
our assumptions and verify:

-   Included coefficients (variables) are significant. For now, only mpg.
-   We havent excluded relevant coefficients that might change significance of
    mpg

\#\#Variable selection

Here comes the hard part. The course makes a point of explaining the process of
variable selection might start

I played around with diagnostic plots, but specially the qq plot cannot mean
anything relevant because we already assume the data is not normal. Its
binomial!

So for now lets stay with our simplest model and try to diagnose it:

```{r fig.height=6}
# autoplot(simplefit_binom_centered_mpg)+
#   theme(text=element_text(size = 5))
```

\#\#Scratch

I mean, what if really mpg has nothing to do with the transmission and we are
missing an effect that is in another variable (and plenty of them have
correlation to both mpg and transmission type).

From the book: Omitting variables results in bias in the coefficients of
interest - unless the regressors are uncorrelated with the omitted ones.

And from the SPLOM plot, we can see mpg is correlated well... with every single
one of the variables!

### Excrutiating Explorations

## The data

The editors have been kind enough to send us a wonderful dataset from the latest
cars in the known universe (1973-1974), wich will allow us to see, FINALLY,
which transmission type is better for our wallet!

You can see it in all its glory bellow:

```{r carstable}
thecars%>%select(mpg,transmission,everything())%>%head()
```

Now that is a lot of variables. We need to see how good (or bad), do those
correlate with each other. A pairs plot is in order as it will guide all of our
work here.

```{r pairwise, echo=FALSE, fig.height=6, fig.width=8}

ggpairs(data = thecars%>%select(-Model),
        ggplot2::aes(colour=transmission),
        upper=list(continuous=wrap("cor",size=2)),
        diag=list(continuous=wrap("diagAxis",labelSize=2,gridLabelSize=2))
        )+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 0))
```

So there we can see a lot of correlated variables. So, what might be interesting
here? For one, most of the variable exhibit some sort of correlation, ranging
from weak to strong, to Miles per Gallon.

In modeling, lets consider what it is we want to answer. One way would be to
attempt to use mpg as the output variables and see if transmission type changes
the average mpg value. It is stated in the problem page: "You work for Motor
Trend, a magazine about the automobile industry. Looking at a data set of a
collection of cars, they are interested in exploring the relationship between a
set of variables and miles per gallon (MPG) (***outcome***)"

So we begin by exploring...

FINAL:

```{r}
lm(mpg~factor(vs)+transmission+wt+disp,thecars)%>%autoplot(which=1:6)
```

### Linear modeling of mpg as a function of transmission

So lets explore first the simplest model and build up from there. The basic
check can be seen in the above SPLOM plot, but lets zoom in the mpg vs.
transmission relationship:

```{r simplest_exploration}
means<-thecars%>%
  group_by(transmission)%>%
  summarize(mean_mpg=mean(mpg))
mean_of_manual<-means[means$transmission=="M",]$mean_mpg
mean_of_automatic<-means[means$transmission=="A",]$mean_mpg
thecars%>%
  ggplot(aes(x=transmission,y=mpg))+
  geom_point(aes(color=transmission))+
  geom_text(data=means,
            aes(x=transmission,y=mean_mpg,color=transmission,
                label=round(mean_mpg,3)),
            nudge_x = -0.2,
            nudge_y = 0.5,
            show.legend = FALSE)+
  geom_hline(data=means,
             aes(yintercept=mean_mpg,color=transmission),
            show.legend = FALSE)
```

Welp. It is clear that Automatic cars have a mean mpg less than the Manual
transmission cars. But are we sure? Like, statistically "sure"? Most certaintly
not. Lets keep exploring this relationship by modeling:
$$MPG=B_0+B_1(transmission); transmission=M:1|A:0$$. Remember then, Manual is 1,
the level of refference and Automatic is 0, the "other" level.

```{r}
fit0<-lm(mpg~transmission,data=thecars)
summary(fit0)
confint(fit0)
```

Okay then. So for this data, if it is the case that all the requirements for
inference are met, we have pretty good significance for the mean mpg given
manual or automatic transmissions. However (there is always one of those), that
is one big **IF**.

Lets diagnose this model and check for the LINE[^1] assumptions:

[^1]: Got it from this EXCELENT resource:
    <https://moderndive.com/10-inference-for-regression.html>

1.  **L**inearity of relationship between variables

2.  **I**ndependence of the residuals

3.  **N**ormality of the residuals

4.  **E**quality of variance of the residuals

Each of those can be assessed by diagnostic plots. Now you may or may not know
this, but the **ggfortify** package can do that and more for us, as it brings
with it cool "autoplot" functions that output ggplot2 plots for diagnosing
models, much like the base R plot().

```{r simplest_diagnostics, fig.width=6,fig.height=6}
autoplot(fit0, which=1:6,label.repel=TRUE)
```

So lets evaluate, reminding that the Automatic group comes first, because it
gets coded as 0:

-   **L**inearity of relationship between variables:

They do seem linear. That is, transmission is binary and the mean within groups
show a linear relationship with mpg.

-   **I**ndependence of the residuals

They are independent due to the way the data is given to us: no car has a
particular relationship with another (except maybe they pay to appear in the
magazzine). We can assume this.

-   **N**ormality of the residuals

The QQ plot shows normality

-   **E**quality of variance of the residuals

The scale-location plot is our friend here and it is saying: no, friend,
variance between the two groups is NOT the same. You need to werk!

And werk we shall...

### Getting rid of heteroskedacity

So there must be some variable that accounts for that variance difference. That
is, removing its effect from the outcome should yield equal variance between the
two groups, so that we might get a precise estimate for the difference in means
per transmission, keeping other variables under control.

Moreover, the scale location plot and the cooks distance vs leverage plot
highlight the worst three offenders. Just to keep ourselves in check, lets look
at them:

```{r fig.width=6}
thecars[c(20,31,29),]
```

So okay. So far so good. These three are some "special" cars in this dataset,
but its not like we see an evident encoding error in mpg so all this leaves us
is to think about what other variables might impact mpg that we aren't taking
into account with just the transmission.

Look back at the SPLOM plot. You see it shows correlations within groups M and
A. We might be looking for a variable that correlates strongly mpg in the manual
group, but not as much in the Automatic group.

One of them is weight, which is in units "1000lbs". It shows -0.909 correlation
with mpg within Manual, but -0.768 within Automatic. Now its not like im just
hunting. It does make a lot of sense that heavier cars would have lesser mpg
regardless of transmission type.

So lets try another model with this variable included:

```{r}
fit1<-update(fit0, mpg~transmission+wt)
summary(fit1)
confint(fit1)
```

Huh. Will you look at that? This knocked the A group's significance quite a bit
but the intercept, which is for manual cars, remains signifficant.

To put it in human, if this is OKAY, then: "For manual transmission cars, per
each ton of weight, average mpg goes down by 7 miles per gallon. For automatic
transmissions, we really dont really, really know whats up".

Which is kind of anti-climatic and I do like working for a magazzine in the
70's. So we gotta werk MOAR. And MOAR we shall Werk.

Maybe we can look at the anova:

```{r}
anova(fit0,fit1)
```

Anova does love this new variable here.

Lets diagnose this new model:

```{r fig.width=8,fig.height=8}
autoplot(fit1,which=1:6)
```

Inclussion of weight really throws things off. We now have a good scale-location
plot, which means (i think) weight did correct the difference in variances, but
the fitted vs. residuals show a clear curve. This is not nice.

So what cars are those in the extremes?

```{r}
thecars[c(17,18,20),]%>%
  select(Model, mpg,transmission,wt,everything())
```

Okay. Well not much there. Now, what other variable might impact mpg? Well if
weight is taken into account and transmission is taken as well, perhaps the
cylinders? I mean, regardless of weight or keeping weight and transmission
constant, more cylinders mean more gas. So, lets do it for that:

```{r}
fit2<-update(fit1,mpg~transmission+wt+factor(cyl))
summary(fit2)
confint(fit2)
```

Sheesh. Everything is signifficant except having an automatic transmission!
grrr. Anova this:

```{r}
anova(fit0,fit1,fit2)
```

This thing likes the cylinders. Maaan... :/.

So cylinders, weight and having a manual transmission are really strong
indicators for the average mpg of car car in this dataset.

Lets diagnose this:

```{r}
autoplot(fit2,which=1:6)
```

Ok so it looks good but its not really what my boss is going to want to publish.
So I did a bit of investigative reporting and found [this
definition](https://www.autolist.com/guides/axle-ratio) for the Rear
Differential Axle Ratio which is recorded for each vehicle in the **drat**
variable:

> "A vehicle's axle ratio is the number of revolutions its driveshaft must make
> to produce one rotation of the axle. Unlike in karts, where wheels connected
> by an axle must rotate at the same speed, most modern vehicles have elaborate
> gear trains on their rear axle to allow the wheels to rotate at different
> rates so they can maintain traction while they turn. The axle ratio for most
> trucks is between 3 and 4, meaning it takes three to four turns of the
> driveshaft to turn the axle once. That is why the ratio is often expressed as
> a number over one, such as 4.10:1...
>
> ...A truck's axle ratio tells you how much torque and towing capacity it can
> produce, and it's an essential indicator of the vehicle's fuel economy. **"**

Okay then. So drat might be one of those "proxy" variables for overall
efficiency of design, since it tells us how fast a motor has to work per
revolution of the rear axle... err... or not...

The first thing here is that its a measure of "strength". That is, how much
power can your car really deliver in its rear axle. As such, it seems to be a
measure of importance more for big cars than for smaller cars. Lets just look at
it by transmission type and also lets look at the car models. Maybe thats what
we aint modeling: if the vehicle is a truck, its mpg will be different from just
a car, regardless of cylinders or transmission, although weight would be another
indicator.

Lets explore this:

```{r}
mpg_vs_drat<-thecars%>%
  ggplot(aes(y=mpg,x=drat,color=wt))+
  geom_point()+
  geom_smooth()
drat_vs_weight<-thecars%>%
  ggplot(aes(y=drat,x=wt,color=mpg))+
  geom_point()+
  geom_smooth()
gridExtra::grid.arrange(mpg_vs_drat,drat_vs_weight)%>%print()
```

Aha! So the more miles per gallon, the more drat, the more drat, the less
weight. Lets see if this helps:

```{r}
fit3<-update(fit2,mpg ~ transmission + wt + factor(cyl)+drat)
summary(fit3)
confint(fit3)
anova(fit0,fit1,fit2,fit3)
```

Now looking back at the correlations in the SPLOM, we can see that drat is the
least correlated variable to mpg in any and all groups.
